---
title: "Group 4 Project 1: Interim Report"
author: Paul Anderson, Michael Barton, Kevin Kincaid
format: pdf
---

```{r}
#| label: document settings
#| include: false
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE)
```

```{r}
#| label: load in packages
#| include: false
library(tidyverse)
library(survey)
library(car)
library(stringr)
library(knitr)
library(kableExtra)
```

## Introduction: Can income be predicted from selected ACS Data

In this interim report, we use individual-level data for Oregon residents from the 2020–2024 American Community Survey (ACS) 5-year Public Use Microdata Sample (1). Exploratory analyses were conducted to evaluate each proposed research question, leading us to focus on the question, “What factors are most impactful in predicting income?” Exploratory analysis is used to assess the distribution of income and its relationships with key demographic and socioeconomic variables. Based on the observed patterns, a full multiple linear regression model is specified, and ANOVA-based comparisons are used to identify an appropriate reduced model and evaluate its performance. We then discuss model justification and potential next steps.

```{r}
#| label: install-deps
#| include: false
# reticulate::virtualenv_create("r-reticulate")
# reticulate::virtualenv_install("r-reticulate", c("requests", "pandas"))
```

```{r}
#| label: add-reticulate
#| include: false
# file.edit("~/.Rprofile")
# add in this 
# Sys.setenv(RETICULATE_PYTHON_ENV = "r-reticulate")
```

```{r}
#| label: optional-pyrestart
#| include: false
# run the following restart the python session if needed
# reticulate::py_run_string("import gc; gc.collect()")
```

```{python}
#| label: fetch-metadata
#| include: false # removed from our output
#| cache: false

import requests
import json
from pathlib import Path

METADATA_PATH = Path("data/pums_variables_metadata.json")

variables = [ # added POBP, JWTRNS, NRC
    "WAGP", "PINCP", "PERNP", "POVPIP", "HICOV", "ESR",
    "AGEP", "SEX", "RAC1P", "HISP", "CIT", "YOEP", "MAR", "RELSHIPP",
    "POBP", "SCHL", "FOD1P", "SCH", "COW", "WKHP", "WKWN", "NAICSP", "OCCP", 
    "WRK", "JWMNP", "JWTRNS", "PUMA", "MIG", "MIGSP", "POWSP",
    "DIS", "DPHY", "DREM", "DEAR", "DEYE","HINCP", "HHT", "TEN", "VEH",
    "NRC", "RETP", "SSP", "SSIP", "PAP", "SEMP",
    "PWGTP", "ADJINC",
]

BASE_URL = "https://api.census.gov/data/2024/acs/acs1/pums/variables/{}.json"

if METADATA_PATH.exists():
    print(f"Loading metadata from cache: {METADATA_PATH}")
    with open(METADATA_PATH) as f:
        metadata = json.load(f)
else:
    print("Fetching variable metadata from Census API...")
    METADATA_PATH.parent.mkdir(parents=True, exist_ok=True)
    metadata = {}

    for var in variables:
        response = requests.get(BASE_URL.format(var), timeout=30)
        if response.status_code == 200:
            metadata[var] = response.json()
            print(f"  OK: {var}")
        else:
            print(f"  FAILED: {var} — {response.status_code}")

    with open(METADATA_PATH, "w") as f:
        json.dump(metadata, f, indent=2)

    print(f"\nSaved metadata for {len(metadata)} variables to {METADATA_PATH}")

# Preview one variable
print(json.dumps(metadata.get("WAGP", {}), indent=2))
```

```{python}
#| label: fetch-pums
#| include: false
#| cache: false

# --- Scrub Start ---

import requests
import pandas as pd
from pathlib import Path

API_KEY = "a8f34453c39e419470f405353250f662a508d821"
YEAR = 2024
STATE = "41"
CACHE_PATH = Path("data/pums_oregon_2024.csv")

variables = [
    # Outcomes
    "WAGP", "PINCP", "PERNP", "POVPIP", "HICOV", "ESR",
    # Core demographics
    "AGEP", "SEX", "RAC1P", "HISP", "CIT", "YOEP", "MAR", "RELSHIPP",
    "POBP",                                                       # added POBP
    # Education
    "SCHL", "FOD1P", "SCH",
    # Labor market
    "COW", "WKHP", "WKWN", "NAICSP", "OCCP", "WRK", "JWMNP", "JWTRNS", # added
    # Geography
    "PUMA", "MIG", "MIGSP", "POWSP",
    # Disability & health
    "DIS", "DPHY", "DREM", "DEAR", "DEYE",
    # Household context
    "HINCP", "HHT", "TEN", "VEH", "NRC",                            # added NRC
    # Income components
    "RETP", "SSP", "SSIP", "PAP", "SEMP",
    # Weights & adjustments
    "PWGTP", "ADJINC","NP",
]

if CACHE_PATH.exists():
    print(f"Loading from local cache: {CACHE_PATH}")
    pums = pd.read_csv(CACHE_PATH, low_memory=False)
    print(f"Loaded: {len(pums):,} rows, {pums.shape[1]} columns")
else:
    print(f"Cache not found — fetching from Census API...")
    CACHE_PATH.parent.mkdir(parents=True, exist_ok=True)
    url = (
        f"https://api.census.gov/data/{YEAR}/acs/acs5/pums"
        f"?get={','.join(variables)}"
        f"&for=state:{STATE}"
        f"&key={API_KEY}"
    )
    response = requests.get(url, timeout=60)
    print(f"Status code: {response.status_code}")
    if response.status_code == 200:
        data = response.json()
        pums = pd.DataFrame(data[1:], columns=data[0])
        pums.to_csv(CACHE_PATH, index=False)
        print(f"Saved to {CACHE_PATH}: {len(pums):,} rows, {pums.shape[1]} columns")
    else:
        print(f"Error: {response.text[:500]}")
```

```{r}
#| label: load-pums
#| include: false

pums <- read_csv("data/pums_oregon_2024.csv", show_col_types = FALSE)

#glimpse(pums)
```

```{r}
#| label: scrub-pums
#| echo: false
#| cache: true

pums_clean <- pums |>

  # --- Remove group quarters records ---
  filter(HINCP != -60000) |>

  # --- Apply income adjustment factor ---
  mutate(
    across(c(WAGP, PINCP, PERNP, HINCP, RETP, SSP, SSIP, PAP, SEMP),
           ~ .x * ADJINC)
  ) |>

  # --- Recode all N/A sentinels to NA ---
  mutate(
    # Income vars: age-based N/A
    WAGP   = na_if(WAGP,  -1 * ADJINC),
    PERNP  = na_if(PERNP, -10001 * ADJINC),
    PINCP  = na_if(PINCP, -19999 * ADJINC),
    RETP   = na_if(RETP,  -1 * ADJINC),
    SSP    = na_if(SSP,   -1 * ADJINC),
    SSIP   = na_if(SSIP,  -1 * ADJINC),
    PAP    = na_if(PAP,   -1 * ADJINC),
    SEMP   = na_if(SEMP,  -10001 * ADJINC),
    # Labor market: 0 = N/A (under 16 or did not work)
    WKHP   = na_if(WKHP, 0),
    WKWN   = na_if(WKWN, 0),
    JWMNP  = na_if(JWMNP, 0),
    # Other sentinels
    POVPIP = na_if(POVPIP, -1),
    ESR    = na_if(ESR, 0),
    YOEP   = na_if(YOEP, 1937),
    HHT    = na_if(HHT, 0),
    TEN    = na_if(TEN, 0),
    VEH    = na_if(VEH, -1),
    NRC    = na_if(NRC, -1),
    DPHY   = na_if(DPHY, 0),
    DREM   = na_if(DREM, 0),
    NP     = na_if(NP, 0),
    # Labor class and work status: 0 = N/A
    COW    = na_if(COW, 0),
    WRK    = na_if(WRK, 0),
  ) |>

  # --- Recode binary vars to logical (1=Yes, 2=No) ---
  mutate(
    across(c(DPHY, DREM, DEAR, DEYE, DIS), ~ .x == 1),
    HICOV = HICOV == 1,
  ) |>

  # --- Recode categorical vars as factors with labels from metadata ---
  mutate(
    SEX = factor(SEX, levels = c(1, 2),
                 labels = c("Male", "Female")),

    RAC1P = factor(RAC1P, levels = 1:9,
                   labels = c("White alone", "Black or African American alone",
                              "American Indian alone", "Alaska Native alone",
                              "AIAN tribes specified", "Asian alone",
                              "Native Hawaiian and Other Pacific Islander alone",
                              "Some other race alone", "Two or More Races")),

    HISP_binary = factor(if_else(HISP == "01", "Not Hispanic", "Hispanic")),

    CIT = factor(CIT, levels = 1:5,
                 labels = c("Born in US", "Born in PR/territories",
                            "Born abroad of US citizen parent",
                            "Naturalized citizen", "Not a US citizen")),

    MAR = factor(MAR, levels = 1:5,
                 labels = c("Married", "Widowed", "Divorced",
                            "Separated", "Never married")),

    ESR = factor(ESR, levels = 1:6,
                 labels = c("Civilian employed at work",
                            "Civilian employed not at work",
                            "Unemployed",
                            "Armed Forces at work",
                            "Armed Forces not at work",
                            "Not in labor force")),

    SCH = factor(SCH, levels = 1:3,
                 labels = c("Not enrolled", "Public school", "Private school")),

    COW = factor(COW, levels = 1:9,
                 labels = c("Private for-profit", "Private nonprofit",
                            "Local govt", "State govt", "Federal govt",
                            "Self-emp incorporated", "Self-emp not incorporated",
                            "Working without pay", "Unemployed last 5 years")),

    MIG = factor(MIG, levels = 1:3,
                 labels = c("Same house", "Outside US", "Different house in US")),

    HHT = factor(HHT, levels = 1:7,
                 labels = c("Married couple household",
                            "Other family: male householder no spouse",
                            "Other family: female householder no spouse",
                            "Nonfamily: male householder living alone",
                            "Nonfamily: male householder not living alone",
                            "Nonfamily: female householder living alone",
                            "Nonfamily: female householder not living alone")),

    TEN = factor(TEN, levels = 1:4,
                 labels = c("Owned with mortgage", "Owned free and clear",
                            "Rented", "Occupied without payment")),

    WRK = factor(WRK, levels = 1:2,
                 labels = c("Worked last week", "Did not work last week")),

    JWTRNS = factor(JWTRNS, levels = 0:12,
                    labels = c("N/A", "Car truck or van", "Bus",
                               "Subway or elevated rail",
                               "Long-distance train or commuter rail",
                               "Light rail streetcar or trolley",
                               "Ferryboat", "Taxi or ride-hailing",
                               "Motorcycle", "Bicycle", "Walked",
                               "Worked from home", "Other method")),

    POBP_binary = factor(case_when(
      as.numeric(POBP) <= 56  ~ "Born in US",
      as.numeric(POBP) == 72  ~ "Born in PR/territories",
      TRUE                    ~ "Foreign born"
    )),
  ) |>

  # --- Drop redundant/replaced columns ---
  select(-ADJINC, -HISP, -POBP, -state) |>

  # --- Ensure valid person weight ---
  filter(!is.na(PWGTP), PWGTP > 0)

#glimpse(pums_clean)
```


```{r}
#| label: eda-overview
#| echo: false

# --- Note ---
# after data pull and scrub in Python from the API
# the following was a proposed EDA method
# the group did chose to work in a different direction for EDA
# we have left if here for possible future use

library(tidyverse)
library(skimr)

# 1. overall summary
skim(pums_clean)
```

```{r}
#| label: eda-missingness
#| include: false
```

```{r}
#| label: eda-income-distributions
#| include: false

# Wages among earners only

# Log wages

# Total personal income
```

```{r}
#| label: eda-outcome-summaries
#| include: false

# Q1/Q3: wage statistics by key groups

# Q2: health insurance coverage rate

# Q2: uninsured rate by citizenship
```

```{r}
#| label: eda-categorical-distributions
#| include: false
```

```{r}
#| label: eda-continuous-distributions
#| include: false
```

```{r}
#| label: eda-correlations
#| include: false

# --- End EDA proposal ---

library(corrr)
```

## Exploratory Data Analysis

We begin by exploring the relationship between income and our key predictors: age, education, gender, race, location and number of household members. 

```{r}
#| label: eda-histograms
#| echo: false
hist(pums_clean$PINCP)
hist(log(pums_clean$PINCP[pums_clean$PINCP > 0]), # only positive earners
     main = "Log Income",
     xlab = "Log PINCP")
```

```{r}
#| label: eda-incomevage
#| echo: false
ggplot(pums_clean, aes(x = AGEP, y = PINCP)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess") +
  labs(title = "Income vs Age")
```

The above summary statistics indicate that income (PINCP) is highly right-skewed, with a small number of individuals earning substantially more than the median. Because of this skewness, we will consider applying a log transformation to the income response when evaluating model fit for the final report. The scatterplot of income versus age suggests a nonlinear relationship: income tends to increase with age early in life, plateau in middle age, and decline slightly at older ages. This pattern suggests the inclusion of a quadratic age term in the regression model.

Missingness evaluation showed income variables are structurally missing for children and non-workers, models will subset to relevant populations rather than imputing values.

```{r}
#| label: eda-incomeved
#| echo: false
ggplot(pums_clean, aes(x = SCHL, y = PINCP)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess") +
  labs(title = "Income vs Education", x = "Years of School (Education)", y = "Income")
```

```{r}
#| label: eda-incomevgender
#| echo: false
ggplot(pums_clean, aes(x = (SEX), y = PINCP)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 200000)) +
  scale_x_discrete(labels = c("Male", "Female")) +
  labs(
    title = "Income by Gender",
    x = "Gender",
    y = "Income"
  ) +
  theme_minimal()
```

```{r}
#| label: eda-incomevrace
#| echo: false
ggplot(pums_clean, aes(x = factor(RAC1P), y = PINCP)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 200000)) +
  labs(
    title = "Income by Race",
    x = "Race",
    y = "Income"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

As shown above, education shows a strong positive association with income, with higher educational attainment corresponding to higher earnings. Boxplots of income by gender and race reveal noticeable differences in median income across groups, suggesting these variables should be included as categorical predictors. Overall, the exploratory analysis indicates that income is associated with multiple demographic and socioeconomic variables, and that nonlinear effects (particularly for age) may be important.Overall, the exploratory analysis indicates that income is associated with multiple demographic and socioeconomic variables, and that nonlinear effects (particularly for age) may be important.

### EDA extension Question 2
We use the full Oregon sample from pums_clean (all ages and labor force statuses) to capture coverage gaps among non-citizens, including those outside the workforce. After dropping missing values for insurance (HICOV), income (POVPIP), and citizenship (CIT), the final sample includes 197,991 residents.

The outcome is being uninsured (1 = uninsured, 0 = insured). Income is measured with POVPIP (percent of the poverty line), aligning with Medicaid and ACA thresholds. Education (SCHL) is grouped into six levels, and employment (ESR_group) into five categories to separate labor force from citizenship effects. Household size (NP), full citizenship categories (CIT), and controls for age, race, and Hispanic origin are also included.

The exploratory analysis reveals a stark and consistent coverage gap by citizenship status. Non-citizens in Oregon are uninsured at roughly 26% — about 21 percentage points higher than US-born residents, whose rate sits between 4–9% across all citizen categories (Figure 1). Income matters: uninsured residents are concentrated below 200% of the poverty line, and POVPIP clearly tracks coverage status (Figure 2). But income tells only part of the story — uninsured individuals appear at every income level, particularly in the 200–400% range, suggesting other forces are at work. Figure 3 makes this plain. Even among the employed, where employer-sponsored coverage is most accessible, non-citizens are uninsured at ~27% compared to ~6% for US-born workers — a gap of over 20 percentage points within the same employment category. Unemployed non-citizens face compounding disadvantage at ~42%, excluded simultaneously from employer coverage and from many public programs available to citizens in identical economic circumstances. The pattern holds across income levels, employment groups, and labor force status: citizenship status appears to independently shape access to health insurance in Oregon, over and above socioeconomic position. The regression models that follow test whether this gap persists after controlling for income, education, employment, household size, age, race, and Hispanic origin.

```{r}
#| label: q2-occp-lookup
#| include: false
#| cache: true

# build OCCP group lookup on distinct codes only
occp_groups <- pums_clean |>
  filter(OCCP != "N", !is.na(OCCP)) |>
  distinct(OCCP) |>
  mutate(
    OCCP_group = case_when(
      between(suppressWarnings(as.numeric(OCCP)), 10,   960)  ~ "High coverage",
      between(suppressWarnings(as.numeric(OCCP)), 1005, 2555) ~ "High coverage",
      between(suppressWarnings(as.numeric(OCCP)), 3000, 3550) ~ "High coverage",
      between(suppressWarnings(as.numeric(OCCP)), 2600, 2920) ~ "Variable coverage",
      between(suppressWarnings(as.numeric(OCCP)), 3600, 3655) ~ "Variable coverage",
      between(suppressWarnings(as.numeric(OCCP)), 4330, 5940) ~ "Variable coverage",
      between(suppressWarnings(as.numeric(OCCP)), 4000, 4255) ~ "Low coverage",
      between(suppressWarnings(as.numeric(OCCP)), 6005, 6130) ~ "Low coverage",
      between(suppressWarnings(as.numeric(OCCP)), 6200, 8990) ~ "Low coverage",
      between(suppressWarnings(as.numeric(OCCP)), 9005, 9760) ~ "Low coverage",
      between(suppressWarnings(as.numeric(OCCP)), 3700, 3960) ~ "Military & Protective",
      between(suppressWarnings(as.numeric(OCCP)), 9800, 9830) ~ "Military & Protective",
      TRUE                                                     ~ "Variable coverage"
    ),
    OCCP_group = factor(OCCP_group,
                        levels = c("High coverage", "Variable coverage",
                                   "Low coverage", "Military & Protective"))
  )

# check
cat("Distinct OCCP codes mapped:", nrow(occp_groups), "\n")
count(occp_groups, OCCP_group)
```
```{r}
#| label: q2-prep
#| include: false
#| cache: false

pums_q2 <- pums_clean |>
  left_join(occp_groups, by = "OCCP") |>
  mutate(
    # response: uninsured
    UNINSURED = as.integer(!HICOV),

    # education collapsed to degree attainment groups
    SCHL_group = case_when(
      SCHL <= 15           ~ "Less than high school",
      SCHL %in% c(16, 17) ~ "High school or GED",
      SCHL %in% c(18, 19) ~ "Some college",
      SCHL == 20           ~ "Associate's degree",
      SCHL == 21           ~ "Bachelor's degree",
      SCHL >= 22           ~ "Graduate degree"
    ),
    SCHL_group = factor(SCHL_group,
                        levels = c("Less than high school", "High school or GED",
                                   "Some college", "Associate's degree",
                                   "Bachelor's degree", "Graduate degree")),

    # employment status collapsed
    # separates labor force status from citizenship for cleaner interpretation
    ESR_group = case_when(
      is.na(ESR)                                          ~ "Under 16",
      ESR == "Civilian employed at work"                  ~ "Employed",
      ESR == "Civilian employed not at work"              ~ "Employed",
      ESR == "Unemployed"                                 ~ "Unemployed",
      ESR %in% c("Armed Forces at work",
                 "Armed Forces not at work")              ~ "Armed Forces",
      ESR == "Not in labor force"                         ~ "Not in labor force"
    ),
    ESR_group = factor(ESR_group,
                       levels = c("Employed", "Unemployed",
                                  "Not in labor force", "Armed Forces",
                                  "Under 16")),

    # age groups for descriptive EDA
    AGE_group = case_when(
      AGEP < 18              ~ "Under 18",
      between(AGEP, 18, 25)  ~ "18-25",
      between(AGEP, 26, 34)  ~ "26-34",
      between(AGEP, 35, 44)  ~ "35-44",
      between(AGEP, 45, 54)  ~ "45-54",
      between(AGEP, 55, 64)  ~ "55-64",
      AGEP >= 65             ~ "65+"
    ),
    AGE_group = factor(AGE_group,
                       levels = c("Under 18", "18-25", "26-34", "35-44",
                                  "45-54", "55-64", "65+"))
  ) |>
  filter(!is.na(HICOV), !is.na(POVPIP), !is.na(CIT))

cat("Q2 analytic sample:", nrow(pums_q2), "residents\n")
cat("Unweighted uninsured rate:",
    round(mean(pums_q2$UNINSURED) * 100, 1), "%\n")
```
```{r}
#| label: q2-survey-design
#| include: false
#| cache: false

# define survey design
svy_q2 <- svydesign(
  ids     = ~PUMA,
  weights = ~PWGTP,
  data    = pums_q2,
  nest    = TRUE
)

cat("Weighted population uninsured rate:",
    round(svymean(~UNINSURED, svy_q2)[[1]] * 100, 1), "%\n\n")
```
```{r}
#| label: q2-theme-setup
#| echo: false
#| cache: false

library(ggplot2)
library(dplyr)
library(stringr)

q2_theme <- theme_classic() +
  theme(
    plot.title    = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    axis.text.y   = element_text(size = 8),
    axis.text.x   = element_text(size = 8),
    plot.margin   = margin(10, 30, 10, 10)
  )
```


```{r}
#| label: q2-eda-citizenship
#| echo: false
#| cache: true

# uninsured rate by citizenship
# visualize citizenship gap
svyby(~UNINSURED, ~CIT, svy_q2, svymean) |>
  mutate(
    pct_uninsured = UNINSURED * 100,
    ci_lower      = (UNINSURED - 1.96 * se) * 100,
    ci_upper      = (UNINSURED + 1.96 * se) * 100,
    CIT           = reorder(CIT, pct_uninsured)
  ) |>
  ggplot(aes(x = CIT, y = pct_uninsured,
             ymin = ci_lower, ymax = ci_upper)) +
  geom_col(fill = "gray", alpha = 0.8) +
  geom_errorbar(width = 0.2) +
  coord_flip() +
  labs(
    title    = "Uninsured Rate by Citizenship Status (weighted)",
    subtitle = "Oregon residents, 2020-2024 ACS 5-Year PUMS",
    x        = NULL,
    y        = "% Uninsured"
  ) +
  theme_classic()
```

```{r}
#| label: q2-fig-povpip
#| echo: false
#| cache: true
#| fig-cap: "Distribution of income-to-poverty ratio by insurance status. Oregon residents, 2020-2024 ACS 5-Year PUMS (unweighted)."

pums_q2 |>
  filter(!is.na(POVPIP)) |>
  ggplot(aes(x = POVPIP,
             fill = factor(UNINSURED, labels = c("Insured", "Uninsured")))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Insured" = "#B0B0B0", "Uninsured" = "#E8A0A0")) +
  labs(
    title    = "Income-to-Poverty Ratio by Insurance Status",
    subtitle = "Oregon residents, 2020-2024 ACS 5-Year PUMS (unweighted)",
    x        = "POVPIP (income as % of poverty line)",
    y        = "Density",
    fill     = NULL
  ) +
  q2_theme +
  theme(legend.position = "bottom")
```

```{r}
#| label: q2-fig-cit-employment
#| echo: false
#| cache: true
#| fig-cap: "Uninsured rate by citizenship status and employment group. Oregon residents, 2020-2024 ACS 5-Year PUMS (weighted). Armed Forces excluded due to small cell sizes."

svyby(~UNINSURED, ~CIT + ESR_group, svy_q2, svymean) |>
  mutate(pct_uninsured = UNINSURED * 100) |>
  filter(ESR_group != "Armed Forces") |>
  mutate(
    CIT       = str_wrap(as.character(CIT), width = 25),
    ESR_group = factor(ESR_group,
                       levels = c("Employed", "Unemployed",
                                  "Not in labor force", "Under 16"))
  ) |>
  ggplot(aes(x = reorder(CIT, pct_uninsured),
             y = pct_uninsured, fill = ESR_group)) +
  geom_col(position = "dodge", alpha = 0.9) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Employed"           = "#B0B0B0",
      "Unemployed"         = "#E8A0A0",
      "Not in labor force" = "#8BAFC4",
      "Under 16"           = "#A8C4A0"
    )
  ) +
  labs(
    title    = "Uninsured Rate by Citizenship and Employment Status",
    subtitle = "Oregon residents, 2020-2024 ACS 5-Year PUMS (weighted)",
    x        = NULL,
    y        = "% Uninsured",
    fill     = "Employment status"
  ) +
  q2_theme +
  theme(
    legend.position = "bottom",
    legend.title    = element_text(size = 9),
    legend.text     = element_text(size = 8)
  ) +
  guides(fill = guide_legend(nrow = 2))
```


## Modeling: Preliminary Models

Based on our analysis of the data, we fit multiple linear regression models to assess how well income can be predicted from the selected variables. The initial full model is:

Income∼Age+Age^2+Education+Gender+Race+Birthplace+Household members

```{r}
#| label: model-1-2-3
#| echo: false
# Urban/Rural is hard to codify in the data, so it may be best to not include it.

library(car)

model1 <- lm(PINCP~AGEP+I(AGEP^2)+SCHL+factor(SEX)+factor(RAC1P)+factor(POBP_binary) + NP, data=pums_clean)
summary(model1)

model2 <- lm(PINCP~AGEP+AGEP^2+SCHL+factor(SEX)+factor(RAC1P), data=pums_clean)
summary(model2)

vif(model1)

model3 <- lm(PINCP~AGEP+SCHL+factor(SEX)+factor(RAC1P), data=pums_clean)
summary(model2)

anova(model2, model3)
```

Income is modeled as a continuous response, and predictors include both continuous variables (age, education, household members) and categorical variables (gender, race, birthplace). A quadratic term for age is included to capture the nonlinear relationship observed in the exploratory analysis.

A reduced model excluding birthplace and household members is also considered due to weak initial evidence of its importance:

Income∼Age+Age^2+Education+Gender+Race

## Model Checking and Selection

```{r}
#| label: model-plots
#| echo: false
plot(model2, which = 1)
plot(model2, which = 2)

#model_log <- lm(log(PINCP + 1) ~ AGEP + I(AGEP^2) + SCHL + factor(SEX) + factor(RAC1P), data = pums_clean)

#lot(model_log, which = 1)
#plot(model_log, which = 2)
```

Model diagnostics were conducted to assess the validity of the linear regression assumptions. Residual plots indicated non-constant variance and right-skewness, which is consistent with the skewed distribution of income. This suggests that a log transformation of income may improve model fit, although the untransformed model is retained for interpretability.

Multicollinearity was assessed using variance inflation factors (VIF). All predictors showed low VIF values (near 1), except for age and age squared, which exhibited moderate collinearity. This is expected due to the polynomial relationship and does not invalidate the model. Centering age can reduce this collinearity but does not change model fit.

Model selection was performed using both hypothesis testing and goodness-of-fit measures. An ANOVA comparison between models with and without the quadratic age term showed that including age squared significantly improves model fit (p < 0.001). In contrast, the inclusion of location (birthplace) did not significantly improve the model and resulted in only a negligible increase in R². Therefore, location was removed from the final model.


The final selected model is:

Income∼Age+Age^2+Education+Gender+Race

## Modeling Question 2
After conducting exploratory data analysis of citizenship, employment status, and their relationship to insurance coverage, we pose the following question:

What is the magnitude of the health insurance coverage gap by citizenship status among Oregon residents, and does this gap persist after adjusting for income, education, employment status, household size, age, race, and Hispanic origin?

### Data and Variables for Question 2
Our data sample uses the full Oregon resident population in pums_clean (n = 197,991), including all ages and labor force statuses. This avoids worker-only restriction, which would underrepresent non-citizens most at risk of uninsurance. Observations missing key variables (HICOV, POVPIP, or CIT) were excluded prior to analysis. 

The response variable is lacking health insurance, coded as a binary integer (1 = uninsured, 0 = insured) by inverting HICOV. It is stored as an integer rather than a logical variable to ensure compatibility with svyby() in the survey package.

Citizenship status (CIT) is the main predictor and is kept in all five categories—U.S.-born, born in Puerto Rico/territories, born abroad to U.S. citizen parents, naturalized citizen, and non-citizen—to capture differences in legal eligibility for coverage.

Income is measured using the income-to-poverty ratio (POVPIP), which is preferred over raw income because eligibility for Medicaid and ACA subsidies is defined relative to the federal poverty line. Education (SCHL) is collapsed into six attainment levels, and employment status (ESR) into five categories to separate labor market effects from citizenship effects.

Household size (NP), age (AGEP), race (RAC1P), and Hispanic origin (HISP_binary) are included as additional controls to adjust for demographic differences across citizenship groups.

### Model Structure
Two survey-weighted logistic regression models are estimated. The reduced model includes only citizenship status and income-to-poverty ratio, providing a baseline estimate of the raw citizenship coverage gap. The full model adds education, employment status, household size, age, race, and Hispanic origin to adjust for socioeconomic and demographic differences. Comparing the two models shows how much of the citizenship gap is explained by income, education, and other characteristics, and how much remains due to structural and legal barriers to coverage.

### Estimation Method
Both models are estimated using survey-weighted logistic regression (svyglm) with a quasibinomial family and logit link. The quasibinomial specification allows for overdispersion by estimating a dispersion parameter rather than fixing it at one, producing more reliable standard errors for survey data. The survey design clusters observations by PUMA and applies person weights (PWGTP) to approximate the PUMS sampling structure.

### Model Comparison and Fit
Model fit is evaluated using quasi-AIC and McFadden’s pseudo $R^2$, both compared to an intercept-only null model. The added contribution of each set of predictors is tested using Wald F-tests (regTermTest()), which is appropriate for survey-weighted models in place of likelihood ratio tests. Multicollinearity is assessed using generalized variance inflation factors (GVIF).

Ridge regression was considered but not used, since the large sample size and relatively small number of predictors make regularization unnecessary, and interpretability of odds ratios is central to the research question on citizenship effects.

### Occupation
Occupation (OCCP) is excluded from the Q2 model even though it is available in the dataset. It contains over 500 SOC codes, and collapsing them into broad categories introduced ambiguity in classification. More importantly, including occupation would restrict the analysis to employed individuals, reintroducing a labor force bias that the full-population design is intended to avoid. Employment status (ESR_group) is retained instead, as it captures labor market attachment without excluding non-workers.

## Question 2 Results
The reduced model (citizenship + income) shows a large raw gap: non-citizens have 6.18 times higher odds of being uninsured than U.S.-born residents (95% CI: 4.99–7.64, p < 0.001). Naturalized citizens show a small, borderline elevation (OR = 1.16, p = 0.071).

In the full model, all major predictor blocks significantly improve fit—education (F = 114.8), employment (F = 289.1), age (F = 311.9), race (F = 14.1), and Hispanic origin (F = 5.75), all p < 0.05—except household size (F = 2.86, p = 0.125). Model fit increases (pseudo R²: 0.075 → 0.135).

The citizenship gap remains large even after adjustment: non-citizens still have 4.64 times higher odds of being uninsured (95% CI: 3.88–5.55), and naturalized citizens also remain at higher risk (OR = 1.31, p = 0.007). This indicates that the gap is only partly explained by socioeconomic factors.

Among controls, higher education is strongly protective (graduate degree OR = 0.207), unemployment increases risk (OR = 1.55), and older age reduces it (OR = 0.981 per year). Certain non-working groups have lower odds due to public coverage programs. Some race effects persist, but should be interpreted cautiously given overlap with citizenship and ethnicity.

## Question 2 Discussion
There is a large and persistent citizenship gap in health insurance coverage in Oregon. In the full survey-weighted logistic model, non-citizens have 4.64 times higher odds of being uninsured than U.S.-born residents (OR = 4.64, 95% CI: 3.88–5.55, p < 0.001) after adjusting for income, education, employment, household size, age, race, and Hispanic origin. This represents about a 25% reduction from the reduced model (OR = 6.18), meaning most of the gap is not explained by observable socioeconomic differences. Naturalized citizens also remain at elevated risk (OR = 1.31, p = 0.007), indicating persistent disadvantage even after legal status changes.

Beyond citizenship, the strongest predictors of being uninsured are:

Education: graduate degree holders are ~80% more likely to have insurance than those without high school completion (OR = 0.207).
Employment: the unemployed are ~55% (OR = 1.55) more likely to be uninsured, although non-working groups do show lower uninsured rates due to public coverage programs (children, Medicare-eligible adults, military coverage).
Age: modest protective effect with increasing age (OR ~ 0.981 per year).
Race/ethnicity: smaller and less stable effects after controls, American Indians remain at higher rates of uninsured (OR = 1.48), while some groups (Asian residents) show lower rates, though estimates are sensitive to multicollinearity.

Key Findings
- The citizenship gap is large and robust to controls, implying it is driven more by structural/legal eligibility constraints than by socioeconomic composition.
- Socioeconomic factors explain only a small part of the disparity, most of the gap persists after adjustment.
- Naturalization does not fully eliminate risk, suggesting residual barriers or post-immigration disadvantages.

Policy implications point to eligibility structure—not income or employment—as the primary driver of coverage inequality. Expanding coverage eligibility regardless of immigration status, reducing waiting periods for immigrants, and improving enrollment access for non-citizens are likely more impactful than labor-market or income-based interventions alone.


```{r}
#| label: q2-full-model
#| echo: false
#| cache: true

# --- Full model ---
# All predictors included
# CIT is focal — all others account for socioeconomic and demographic differences

model_q2_full <- svyglm(
  UNINSURED ~ CIT + POVPIP + SCHL_group + ESR_group + NP + AGEP + RAC1P + HISP_binary,
  design = svy_q2,
  family = quasibinomial(link = "logit")
)

summary(model_q2_full)

# Odds ratios with 95% CI
or_full <- exp(cbind(
  OR    = coef(model_q2_full),
  lower = coef(model_q2_full) - 1.96 * sqrt(diag(vcov(model_q2_full))),
  upper = coef(model_q2_full) + 1.96 * sqrt(diag(vcov(model_q2_full)))
)) |>
  round(3) |>
  as.data.frame() |>
  rownames_to_column("term")

print(or_full)

# Multicollinearity
vif(model_q2_full)
```
```{r}
#| label: q2-reduced-model
#| echo: false
#| cache: true

# --- Reduced model ---
# citizenship + income only

model_q2_reduced <- svyglm(
  UNINSURED ~ CIT + POVPIP,
  design = svy_q2,
  family = quasibinomial(link = "logit")
)

summary(model_q2_reduced)

or_reduced <- exp(cbind(
  OR    = coef(model_q2_reduced),
  lower = coef(model_q2_reduced) - 1.96 * sqrt(diag(vcov(model_q2_reduced))),
  upper = coef(model_q2_reduced) + 1.96 * sqrt(diag(vcov(model_q2_reduced)))
)) |>
  round(3) |>
  as.data.frame() |>
  rownames_to_column("term")

print(or_reduced)
```
```{r}
#| label: q2-model-comparison
#| include: false
#| cache: true

# --- Model comparison ---

# Wald F-tests for each predictor block
cat("--- Wald F-tests: does each block add beyond CIT + POVPIP? ---\n\n")
cat("Education:\n");        print(regTermTest(model_q2_full, ~ SCHL_group))
cat("Employment:\n");       print(regTermTest(model_q2_full, ~ ESR_group))
cat("Household size:\n");   print(regTermTest(model_q2_full, ~ NP))
cat("Age:\n");              print(regTermTest(model_q2_full, ~ AGEP))
cat("Race:\n");             print(regTermTest(model_q2_full, ~ RAC1P))
cat("Hispanic origin:\n");  print(regTermTest(model_q2_full, ~ HISP_binary))

# quasi-AIC
cat("\nQuasi-AIC - Reduced:", AIC(model_q2_reduced)[2],
    "\nQuasi-AIC - Full:   ", AIC(model_q2_full)[2], "\n")

# McFadden pseudo R-squared
null_model <- svyglm(
  UNINSURED ~ 1,
  design    = svy_q2,
  family    = quasibinomial(link = "logit")
)

mcfadden_full    <- 1 - deviance(model_q2_full)    / deviance(null_model)
mcfadden_reduced <- 1 - deviance(model_q2_reduced) / deviance(null_model)

cat("\nMcFadden R^2 - Reduced:", round(mcfadden_reduced, 4),
    "\nMcFadden R^2 - Full:   ", round(mcfadden_full, 4), "\n")

# key result: how much does the citizenship gap shrink after controls?
cat("\n--- Citizenship odds ratios: reduced vs full model ---\n")

cit_terms <- paste0("CIT", levels(pums_q2$CIT)[-1])

tibble(
  citizenship = levels(pums_q2$CIT)[-1],
  OR_reduced  = or_reduced$OR[or_reduced$term %in% cit_terms],
  OR_full     = or_full$OR[or_full$term %in% cit_terms],
  pct_change  = round((OR_full - OR_reduced) / OR_reduced * 100, 1)
) |>
  print()
```

```{r}
#| label: q2-fig-forest
#| echo: false
#| fig-height: 4
#| fig-width: 7

bind_rows(
  or_reduced |>
    filter(str_detect(term, "^CIT")) |>
    mutate(
      model       = "Reduced",
      citizenship = str_remove(term, "^CIT")
    ),
  or_full |>
    filter(str_detect(term, "^CIT")) |>
    mutate(
      model       = "Full",
      citizenship = str_remove(term, "^CIT")
    )
) |>
  mutate(
    citizenship = str_wrap(citizenship, width = 25),
    model       = factor(model, levels = c("Reduced", "Full"))
  ) |>
  ggplot(aes(x = OR, y = reorder(citizenship, OR),
             color = model, shape = model)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray60") +
  geom_errorbar(aes(xmin = lower, xmax = upper),
                width = 0.15, linewidth = 0.7,
                position = position_dodge(width = 0.5)) +
  geom_point(size = 4,
             position = position_dodge(width = 0.5)) +
  scale_color_manual(
    values = c("Reduced" = "#6B6B6B", "Full" = "#C0392B"),
    labels = c("Reduced (CIT + income only)",
               "Full (all predictors)")
  ) +
  scale_shape_manual(
    values = c("Reduced" = 16, "Full" = 17),
    labels = c("Reduced (CIT + income only)",
               "Full (all predictors)")
  ) +
  scale_x_log10(breaks = c(0.5, 1, 2, 3, 5, 7),
                labels = c("0.5", "1", "2", "3", "5", "7")) +
  labs(
    title    = "Odds of Being Uninsured by Citizenship Status",
    subtitle = "Reference: Born in US — Oregon residents, 2020-2024 ACS 5-Year PUMS",
    x        = "Odds Ratio (log scale)",
    y        = NULL,
    color    = "Model",
    shape    = "Model"
  ) +
  q2_theme +
  theme(
    legend.position  = "bottom",
    legend.title     = element_blank(),
    panel.grid.major.x = element_line(color = "gray90", linewidth = 0.4)
  ) +
  guides(color = guide_legend(nrow = 1))
```

```{r}
#| label: q2-results-table
#| echo: false
#| cache: true

# showing change ratio and significance
cit_terms <- c("CITNot a US citizen",
               "CITNaturalized citizen",
               "CITBorn in PR/territories",
               "CITBorn abroad of US citizen parent")

or_reduced <- exp(coef(model_q2_reduced))[cit_terms]
or_full    <- exp(coef(model_q2_full))[cit_terms]
p_full     <- summary(model_q2_full)$coefficients[cit_terms, "Pr(>|t|)"]
pct_change <- round((or_full - or_reduced) / or_reduced * 100, 1)

p_fmt <- ifelse(p_full < 0.001, "< 0.001",
                formatC(p_full, format = "f", digits = 3))

change_fmt <- ifelse(pct_change > 0,
                     paste0("+", pct_change, "%"),
                     paste0(pct_change, "%"))

tibble(
  `Citizenship Status` = c("Not a US citizen",
                            "Naturalized citizen",
                            "Born in PR/territories",
                            "Born abroad of US citizen parent"),
  `OR (Reduced)`       = round(or_reduced, 2),
  `OR (Full)`          = round(or_full, 2),
  `Change`             = change_fmt,
  `p (Full)`           = p_fmt
) |>
  kable(
    caption  = "Citizenship odds ratios, reduced vs full model (ref: Born in US). Change reflects OR shift after adding socioeconomic controls.",
    align    = c("l", "c", "c", "c", "c"),
    booktabs = TRUE
  ) |>
  kable_styling(
    latex_options     = c("striped", "hold_position"),
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    font_size         = 10
  ) |>
  row_spec(1, bold = TRUE)

# --- Model fit summary ---
tibble(
  Metric          = c("McFadden R²", "Quasi-AIC"),
  `Reduced Model` = c("0.075", "81,935"),
  `Full Model`    = c("0.135", "76,687")
) |>
  kable(
    caption  = "Model fit comparison: reduced vs full model.",
    align    = c("l", "c", "c"),
    booktabs = TRUE
  ) |>
  kable_styling(
    latex_options     = c("striped", "hold_position"),
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    font_size         = 10
  )
```

## Conclusions

Our analysis shows that income can be partially predicted using age, education, gender, and race. Education is the strongest predictor, with higher levels of education associated with substantially higher income. Age has a nonlinear relationship with income, with earnings increasing early in life, peaking in mid-career, and declining slightly thereafter. Gender and race also show statistically significant associations with income, even after controlling for other variables.

Despite these significant relationships, the overall explanatory power of the model is modest (R² = 0.12), indicating that we may need to consider other variables as determinants of income. Going forward there are additional factors to consider. One would be exploring additional variables for positive associates with income that could be added to the model. Another option is testing interaction terms. We could also consider whether gender and race are impactful enough that we should really be developing separate models in order to accurately capture the full impact these categorical variables have on all other aspects. 


## Citations

1. U.S. Census Bureau. (2026). 2020–2024 American Community Survey 5-year Public Use Microdata Sample (PUMS). U.S. Department of Commerce. https://www.census.gov/programs-surveys/acs/microdata.html

2. Mason, H., & Wiggins, C. (2010). A Taxonomy of Data Science. (Archived manuscript / teaching document). Accessed on 4/10/2026 https://introdatasci.dlilab.com/pdf/A_Taxonomy_of_Data_Science.pdf

## Appendix

```{r appendix-collect, include=FALSE}
all_chunks <- setdiff(knitr::all_labels(), c("appendix-collect", "appendix"))
```

```{r appendix, ref.label=all_chunks}
#| echo: true
#| eval: false

```